---
title: "R Processing of Empirical Hamlets"
author: "Kate Shih"
date: "2023-11-04"
output: html_document
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This document will guide you how to apply R to generate Empirical Hamlets using OSM building centroids and conduct population estimation for the hamlets.

```{r warning = FALSE, message = FALSE}
#Allow R to load required algorithms
library(sp)
library(sf)
library(dplyr)
library(tidyverse)
library(janitor)
library(extraDistr)
```

## Step 1. Upload your OSM building centroids .shp

After running the codes, another window will be open and direct to the folder where the shapefile is then click "Open".
The file will be loaded.

After loading completely, an output will show the number of OSM buildings points we have.

```{r}
hh_all<-st_read(file.choose())

nrow(hh_all)
```

## Step 2. Clean building points

Because we are targeting "population households" and they are not likely to be labeled with "Name" on the map, we removed points labeled with "Name".

After filtering completely, an output will show the number of clean points we have.

```{r}
hh<-hh_all %>% 
  filter(is.na(name)|is.na(name))%>% 
  dplyr::select(osm_id,longitude,latitude) 

nrow(hh)
```

## Step 3. Transform the spatial data to a general data table

To conduct K-means clustering, we need to extract the spatial coordinates into a plain data table.

```{r}
hh.loc<-as.data.frame(st_coordinates(hh))
```

## Step 4. Apply K-means clustering to building centroids

The core concept of "Empirical Hamlets" is hamlets consist of clustering of households.
To apply K-means clustering, we need to pre-define "K", the number of clusters we would like to generate.
We refer the denominator from the real hamlets in India.
In average, 51 households (14376/282) would compose one hamlet.

"km" is the output of K-means clustering and it contains lists of:

1.  cluster: A vector of integers (from 1:k) indicating the cluster to which each point is allocated.

2.  centers: A matrix of cluster centers.

3.  size: The number of points in each cluster.

*###Setting a seed allows us to get the same result every time we run the code.*

*###A iteration time of 50 should be sufficient to have the algorithm to coverge.*

*###Please be patient. We have quite a huge dataset...*

```{r}
k<-round(nrow(hh.loc)/(14376/282),0)
set.seed(123)
km <-kmeans(hh.loc, k, iter.max = 50, trace = T)
```

## Step 5. Conduct population estimation

Here we applied District population as our reference; therefore, we need to upload district shapefile [adm2.shp] and population csv [Dist_CEN_2021.csv] first.
Next, we intersected (st_join) the building centroids with districts to identify which district the building is located.
Then, we can simulate the household size for each building point based on the mean household size of that district.
The algorithm was randomized under zero-truncated Possion distribution (rtpois).

```{r}
#upload district shapefile
adm2<-st_read(file.choose())

#upload district population data
adm2_pop<-read.csv(file.choose())

#join population to district shapefile
adm2<-left_join(adm2,adm2_pop,join_by(ADM2_PCODE==ADM2_PCODE))

#just to reduce the data table size
adm2<-adm2 %>% 
  dplyr::select(ADM2_EN,ADM2_PCODE,HH_SIZE,TOT_POP)

#add k-means cluster result to the building centroid dataset
hh$kmcluster<-km$cluster

hh_in <- st_join(hh, adm2, join = st_within)

#just a reference table to know how many districts are within scope
count<-count(as_tibble(hh_in), ADM2_PCODE)

#hh_size is the simulated household size
hh_in$hh_size<-vector(length=nrow(hh_in))
for (i in 1:nrow(count)){
  pp.size<-adm2$HH_SIZE[adm2$ADM2_PCODE==count$ADM2_PCODE[i]] #District HOUSEHOLD
  pp.pop<-adm2$TOT_POP[adm2$ADM2_PCODE==count$ADM2_PCODE[i]] #TOTAL POPULATION
  set.seed(10)
  size<-rtpois(count$n[i], pp.pop/pp.size, a = 0, b = Inf)
  hh_in$hh_size[hh_in$ADM2_PCODE==count$ADM2_PCODE[i]]<-size
}

#Consize the 'hh_in' dataset
hh_in<-hh_in %>% 
  dplyr::select(osm_id,longitude,latitude,ADM2_EN,ADM2_PCODE,kmcluster,hh_size)

```

## Step 6. Set up Hamlet Center Dataset

To complete the population esimation of each hamlet, we just need to sum up 'hh_size' of the same cluster from building centroids.
Meanwhile, we will set up a point shapefile [center] as the centroid dataset for our empirical hamlets with variables of household numbers ('size') and population estimates ('pop_est').

```{r}
cen<-as.data.frame(km[["centers"]])
cen$kmcluster<-1:k
cen$size<-
  unlist(km$size)
center<-st_as_sf(cen, coords = c("X","Y"), crs=4326, remove = FALSE)

#just to add which district this hamlet locates
adm2c<-adm2 %>% 
  dplyr::select(ADM2_EN,ADM2_PCODE)
c_in2 <- st_join(center, adm2c, join = st_within)
center$ADM2_EN<-c_in2$ADM2_EN
center$ADM2_PCODE<-c_in2$ADM2_PCODE

#pop_est is the population estimates of our hamlet
center$pop_est<-vector(length=nrow(center))
for (i in 1:k){
  cluster_hh<-hh_in %>% 
    filter(kmcluster==i)
  center$pop_est[center$kmcluster==i]<-sum(cluster_hh$hh_size)
}
```

## Step 7. Set up Building Polygong Dataset

Next, we will upload the building polygon shapefile and then join the variables of cluster ('kmcluster') and household size ('hh_size') for further applications.

```{r}
hhp<-st_read(file.choose())

hhp<-hhp %>% 
  dplyr::select(osm_id) %>% 
  filter(osm_id %in% hh_in$osm_id)

dt<-hh_in %>% 
  dplyr::select(osm_id,kmcluster,ADM2_EN,ADM2_PCODE,hh_size) %>% 
  st_drop_geometry()

hhp<-left_join(hhp,dt,join_by(osm_id==osm_id))
```

## Step 8. Export the results!!!

Now, we can save our results!
We will export 3 shapefiles:

1.  center: the centroids of our empirical hamlets with hamlet sequence ('kmcluser'), hamlet size ('size'), and population estimates ('pop_size').

2.  hh_in: the centroids of building with hamlet sequence ('kmcluser'), household size estimates ('hh_size').

3.  hhp: the building polygon with hamlet sequence ('kmcluser'), household size estimates ('hh_size').

*###The exporting code is a bit trickier than uploading.* *###Please follow the instruction within the codes.*

```{r}
#To avoid OSMID alternation during data writing, we transform the data type
hh_in$osm_id<-as.character(hh_in$osm_id)
hhp$osm_id<-as.character(hhp$osm_id)

#Inside the bracket, please replace your Folder path, and be aware of the direction "/"
setwd("E:/Nepal/Settlement/final")

#Then run these lines directly!
st_write(center,"EmpiricalHamlets_Centroids.shp")
st_write(hh_in,"HamletsBuilding_Centroids.shp")
st_write(hhp,"HamletsBuilding_Polygon.shp")
```
